pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools

### 2020-08-04 bmc_article.tex
4overleaf

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage[square,sort,comma,numbers]{natbib}
\usepackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{graphicx}
\usepackage{url}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1,aff2,aff3},          % id's of addresses, e.g. {aff1,aff2}
   email={pierre-luc.germain@hest.ethz.ch}   % email address
]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
\author[
   addressref={aff1,aff2},
    email={anthony.sonrel@uzh.ch}
]{\fnm{Anthony} \snm{Sonrel}}
\author[
   addressref={aff1,aff2},
   corref={aff1},  % id of corresponding address, if any
   email={mark.robinson@imls.uzh.ch}
]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{
  \orgname{Department of Molecular Life Sciences, University of Zürich}, % university, etc
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}
\address[id=aff2]{
  \orgname{SIB Swiss Institute of Bioinformatics}, % university, etc
  \city{Zürich},
  \cny{Switzerland}
}
\address[id=aff3]{
  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract}

\parttitle{Background} Despite the massive growth of single-cell RNA-sequencing (scRNAseq) and corresponding computational methods, sufficient and up-to-date benchmarks to guide analytical choices are still lacking. Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline. Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis.

\parttitle{Results} Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics. We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering.

\parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline. The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{single-cell RNA sequencing (scRNAseq)}
\kwd{pipeline}
\kwd{clustering}
\kwd{normalization}
\kwd{dimensionality reduction}
\kwd{filtering}
\kwd{benchmark}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of a scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

@@

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately{\color{red}, as it was not meant for this purpose and follows a slightly different flow}). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. {\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15). In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}. {\color{red} We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.}

We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B). Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.

{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

Of note, the optimal method in our evaluation (\textit{maxLikGlobal})} systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 35:

@@

{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}. Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space. These studies, like ours, used datasets with relatively few cells (i.e. fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted. We think that further work will be required to assess its performance on larger datasets.}

>>>>

\documentclass[11pt]{article}

\renewcommand\familydefault{\sfdefault}
\usepackage[left=3cm,right=3cm,top=2cm,bottom=3cm]{geometry}

% graphics, colors & math
\usepackage{graphicx}
\usepackage[dvipsnames,usenames]{xcolor}
\usepackage{amsmath,amssymb, amsfonts,nicefrac,textcomp,icomma,bm}

% links & figures
\usepackage{hyperref}
\usepackage[labelfont=bf,font=small]{caption}
\captionsetup[figure]{font={stretch=1}}
\newcommand{\bcaption}[2]{\caption[#1]{\textbf{#1} #2}}
\usepackage{float}

% cross-referencing
\usepackage{xr}
\makeatletter
\newcommand*{\addFileDependency}[1]{%
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\usepackage{pdfpages}

% line spacing
\usepackage{setspace}
\renewcommand{\baselinestretch}{1.5}
\setlength{\parindent}{2em}

% line numbering
\usepackage{etoolbox,lineno}
\newcommand*{\Section}[1]{\par\nolinenumbers\section*{#1}\linenumbers}

% title layout
\usepackage[affil-it]{authblk}
\renewcommand\Authands{ \& }

\usepackage{lipsum}
\makeatletter
\renewcommand{\maketitle}{\bgroup
\setlength{\parindent}{0pt}\begin{flushleft}
\textbf{\Large{\@title}}\\\vspace{1em}\@author
\end{flushleft}\egroup}
\makeatother

\author[1,2,3,*]{Pierre-Luc Germain}
\author[1,2]{Anthony Sonrel}
\author[1,2,*]{Mark D. Robinson}

\affil[1]{Department of Molecular Life Sciences, University of Z{\"u}rich, Switzerland}
\affil[2]{SIB Swiss Institute of Bioinformatics, Z{\"u}rich, Switzerland}
\affil[3]{D-HEST Institute for Neuroscience, Swiss Federal Institute of Technology (ETH), Z{\"u}rich, Switzerland}
\affil[*]{Correspondence to Pierre-Luc Germain (pierre-luc.germain@hest.ethz.ch) and Mark D. Robinson (mark.robinson@imls.uzh.ch)}

\date{\today}

% code font-style
\newsavebox{\foobox}
\newcommand{\slantbox}[2][0]{\mbox{%
        \sbox{\foobox}{#2}%
        \hskip\wd\foobox
        \pdfsave
        \pdfsetmatrix{1 0 #1 1}%
        \llap{\usebox{\foobox}}%
        \pdfrestore
}}
\newcommand\unslant[2][-.25]{\slantbox[#1]{$#2$}}
\newcommand{\code}[1]{\unslant{\mathit{#1}}}

% citations
\usepackage[style=nature,citestyle=numeric,sorting=none]{biblatex}
\renewcommand*{\bibfont}{\footnotesize}
\addbibresource{refs.bib}

\renewcommand{\cite}[1]{\textcolor{Blue}{$^[$\supercite{#1}$^]$}}
\newcommand{\citet}[1]{\citeauthor{#1}\textcolor{Blue}{$^[$\supercite{#1}$^]$}}

% check marks
\usepackage{pifont}
\newcommand{\cmark}{\textcolor{Green}{\ding{51}}}
\newcommand{\xmark}{\textcolor{Red}{\ding{55}}}

% tables
\usepackage{colortbl}

\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}

\begin{document}

\thispagestyle{empty}
\maketitle

\linenumbers
%\modulolinenumbers[5]

\Section{Abstract}

The massive growth of single-cell RNA-sequencing (scRNAseq) and the methods for its analysis still lack sufficient and up-to-date benchmarks that could guide analytical choices. Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline. Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics. We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering. On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline. The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.

\newpage
\pagenumbering{arabic}

\section*{Background}
Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology \cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}. However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of a scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \cite{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \cite{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \cite{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

@@

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.

We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B). Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model \textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:

@@
